---
title: "Count data processing"
output:
  html_document:
    df_print: paged
    css: ../resources/styles.css
---
```{r setup, include = FALSE}
library(tidyverse)
```

This notebook contains the instructions to process the count matrix output from the STAR-RSEM pipeline.
The outputs from this script are:

* Unnormalized counts with the full species-specific gene set (without orthologs)
* Unnormalized counts with one-to-one orthologs

The counts are normalized with DESeq2 in the server outside this notebook. The normalized counts are then further processed here to get:

* Normalized counts both with species-specific gene set and orthologs
* Per-slice average normalized counts. 
* Z-score transformed normalized counts. The transformation is done across all conditions/slices within a single species.

Input data
```{r}
orthologs <- readRDS("../../datasets/gene_orthologs/Mus-Acomys_OrthoFinder_1to1_orthologs_wSymbols.RData")
unnorm_counts_dir <- "../../datasets/gene_counts/un-normalized_counts/"
star_rsem_counts_paths <- list.files(path = paste0(unnorm_counts_dir, "STAR-RSEM_output/"), full.names = TRUE)
```

#### Functions
```{r}
#Function for processing the unnormalized gene counts
format_unnorm_counts <- function(file_path, output_dir_path = unnorm_counts_dir) {
  file_name <- basename(file_path)
  sample_prefix <- gsub("^(.+?)_gene_counts_.*$", "\\1", file_name)
  
  unnorm_counts <- read.delim(file_path) %>%
    rownames_to_column("Gene_EnsID") %>%
    mutate(Gene_EnsID = str_remove(Gene_EnsID, "gene:"))
  
  write_tsv(unnorm_counts, paste0(output_dir_path, sample_prefix, "_STAR-RSEM_un-normalized.tsv"))
}

# Function for creating the unnormalized counts with 1-1 orthologs
format_ortholog_counts <- function(file_path, ortholog_data = orthologs, output_dir_path = unnorm_counts_dir) {
  
  # Fetch species symbol and file prefix from the file path string
  file_name <- basename(file_path)
  species <- gsub("^([A-Za-z]+)_.*$", "\\1", file_name)
  if (!(species %in% c("MM", "AC"))) {
    stop("Accepted species symbols: MM, AC")
  }
  sample_prefix <- gsub("^(.+?)_STAR-RSEM_.*$", "\\1", file_name)
  
  # Format to ortholog counts according to species
  count_data <- read.delim(file_path)
  if (species == "AC") {
    count_data_orthologs <- count_data %>%
      inner_join(., ortholog_data, by = join_by(Gene_EnsID == Acomys_EnsID)) %>%
      select(-Gene_EnsID) %>%
      rename(Gene_EnsID = Mus_EnsID) %>%
      relocate(Gene_symbol, Gene_EnsID)
  
  } else if (species == "MM") {
    count_data_orthologs <- count_data %>%
      inner_join(., ortholog_data, by = join_by(Gene_EnsID == Mus_EnsID)) %>%
      select(-Acomys_EnsID)
  
  } else {
    stop("Unknown species symbol")
  }
  
  # Save
  write_tsv(count_data_orthologs, 
            paste0(output_dir_path, "un-normalized_counts_1-1orth/", sample_prefix, "_STAR-RSEM_un-normalized_1-1orth.tsv")
            )
}

# Function to combine unnormalized counts
combine_dir_count_files <- function(file_paths, with_orthologs = FALSE) {
  # Read all files into a file list
  file_list <- lapply(file_paths, function(path) {
    read.delim(path) 
    })
  
  # Start with the first file and iteratively join the remaining files
  combined_data <- file_list[[1]]
  for (i in 2:length(file_list)) {
    
    if (with_orthologs){
      combined_data <- inner_join(combined_data, file_list[[i]], by = c("Gene_EnsID", "Gene_symbol"))
    }
    
    if (!with_orthologs) {
      combined_data <- inner_join(combined_data, file_list[[i]], by = c("Gene_EnsID"))
    }
    
  }
  return(combined_data)
}
```

Process the un-normalized counts per replicate
```{r}

# Unnormalized counts without orthologs
for (file_path in star_rsem_counts_paths) {
  format_unnorm_counts(file_path = file_path)
}
rm(file_path)

# Unnormalized counts with orthologs
unnorm_counts_paths <- list.files(path = unnorm_counts_dir, pattern = ".tsv$", full.names = TRUE)
for (file_path in unnorm_counts_paths) {
  format_ortholog_counts(file_path = file_path)
}
rm(file_path)
```

Combine counts per species
```{r}
mus_unnorm_counts <- list.files(path = unnorm_counts_dir, pattern = "^MM", full.names = TRUE) %>%
  combine_dir_count_files() %>%
  saveRDS(paste0(unnorm_counts_dir, "Mus_STAR-RSEM_un-normalized.RData"))
  
mus_unnorm_ortholog_counts <- list.files(path = paste0(unnorm_counts_dir, "un-normalized_counts_1-1orth/"), pattern = "^MM", full.names = TRUE) %>%
  combine_dir_count_files(., with_orthologs = TRUE) %>%
  saveRDS(., paste0(unnorm_counts_dir, "un-normalized_counts_1-1orth/", "Mus_STAR-RSEM_un-normalized_1-1orth.RData"))

acomys_unnorm_counts <- list.files(path = unnorm_counts_dir, pattern = "^AC", full.names = TRUE) %>%
  combine_dir_count_files() %>%
  saveRDS(., paste0(unnorm_counts_dir, "Acomys_STAR-RSEM_un-normalized.RData"))
  
acomys_unnorm_ortholog_counts <- list.files(path = paste0(unnorm_counts_dir, "un-normalized_counts_1-1orth/"), pattern = "^AC", full.names = TRUE) %>%
  combine_dir_count_files(., with_orthologs = TRUE) %>%
  saveRDS(., paste0(unnorm_counts_dir, "un-normalized_counts_1-1orth/", "Acomys_STAR-RSEM_un-normalized_1-1orth.RData"))
```

Work with normalized counts