---
title: "Median of ratios normalization visualization"
author: "Arne Kukkonen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dendextend)
```

Load the raw un-normalized counts
```{r}
input_dir <- "datasets/Un-normalized_counts/"
mm_counts <- readRDS(paste0(input_dir, "MM_Tomoseq_counts_1-1orth_unnorm.RData"))
ac_counts <- readRDS(paste0(input_dir, "AC_Tomoseq_counts_1-1orth_unnorm.RData"))
```

###Subset the sample groups

Functions to subset the sample groups
```{r}
# Function to subset slice intervals
slice_subset <- function(data, slice_numbers) {
    data_sub_slices <- data %>% 
      select(GeneID, matches(paste0("\\.slice(", paste(slice_numbers, collapse="|"), ")$")))
    return(data_sub_slices)
}

# Function to subset counts based on condition and slice numbers
sample_subset <- function(data, cond_pattern, slice_numbers) {
  data_sub <- data %>%
    select(GeneID, matches(cond_pattern)) %>%
    slice_subset(., slice_numbers = slice_numbers)
  return(data_sub)
}
```

Create count subsets for both species
```{r}
# Define slice zone intervals
fibrotic_slices <- c(16:24)
remote_slices <- c(1:13)
sham_slices <- c(1:24)

# Mus samples
mm_fibrotic <- sample_subset(mm_counts, cond_pattern = "MI14D", slice_numbers = fibrotic_slices)
mm_remote <- sample_subset(mm_counts, cond_pattern = "MI14D", slice_numbers = remote_slices)
mm_sham <- sample_subset(mm_counts, cond_pattern = "SHAM", slice_numbers = sham_slices)

# Acomys samples
ac_fibrotic <- sample_subset(ac_counts, cond_pattern = "MI14D", slice_numbers = fibrotic_slices)
ac_remote <- sample_subset(ac_counts, cond_pattern = "MI14D", slice_numbers = remote_slices)
ac_sham <- sample_subset(ac_counts, cond_pattern = "SHAM", slice_numbers = sham_slices)
```

### meta data
Function to create meta data
```{r}
get_meta <- function(count_data) {
  
  sample_names <- count_data %>%
    select(-GeneID) %>%
    colnames()
  
  sample_conditions <- str_extract(sample_names, "(?<=_).*?(?=\\.)")
  sample_species <- str_extract(sample_names, "^[A-Z]{2}")
  
  meta_data <- data.frame(row.names = sample_names, Condition = sample_conditions, Species = sample_species)
  
  # Add the zone annotation based on condition and slice numbers
  meta_data <- meta_data %>%
    mutate(
      Zone = if_else(
        Condition == "MI14D" & between(as.numeric(str_extract(rownames(.), "\\d+$")), 16, 24),
        true = "Fibrotic", false = "Sham"),
      Zone = if_else(
        Condition == "MI14D" & between(as.numeric(str_extract(rownames(.), "\\d+$")), 1, 13),
        true = "Remote", false = Zone),
      )
  
  return(meta_data)
}
```

Create sample metadata for species, experimental condition and the MI zone. 
```{r}
# Combine the dataset
combined_counts <- mm_fibrotic %>%
  # Mus samples
  inner_join(., mm_sham, by = join_by(GeneID)) %>%
  # Acomys samples
  inner_join(., ac_fibrotic, by = join_by(GeneID)) %>%
  inner_join(., ac_sham, by = join_by(GeneID))

meta_data <- get_meta(combined_counts)
```

Normalize the counts
```{r}
# Normalize counts for clustering
cc <- combined_counts %>%
  column_to_rownames("GeneID") %>%
  edgeR::cpm() %>%
  log1p()
  #(function(x) x/1e5)

if(!all.equal(rownames(meta_data), colnames(cc))) {
  stop("row names of meta and column names of counts do not match")
} else {
  print("check passed")
}
```
Create meta data color mapping for the color bar
```{r}
# Create meta data color mapping for the color bar
species_colors <- c("MM" = "#33608C", 
                    "AC" = "#8867A1") 
species_color_vector <- species_colors[meta_data$Species]
zone_colors <- c("Fibrotic" = "#CA3B43",
                 "Remote"  = "#F5B355",
                 "Sham" = "#C7C9CA")
zone_color_vector <- zone_colors[meta_data$Zone]
color_matrix <- cbind(species_color_vector, zone_color_vector)
```

Run clustering on the samples
```{r}
# Calculate distances among samples (columns)
dist_mat <- dist(t(cc), method = "euclidean")

# Perform hierarchical clustering
hc <- hclust(dist_mat, method = "complete")
```

Plot the dendrogram with species
```{r}
# Open PNG device - specify file name and dimensions
png("dendrogram_sham.png", width = 800, height = 650, res = 200)  # You can adjust width/height as needed

# Convert hclust to dendrogram
dend <- as.dendrogram(hc)

# Get species in dendrogram order
variable_ordered <- meta_data$Species[order.dendrogram(dend)]

# Create integer clusters (1 for MM, 2 for AC)
variable_clusters <- as.numeric(factor(variable_ordered))

# Color branches using these clusters
dend <- color_branches(dend,
                      clusters = variable_clusters,  # Integer vector matching leaf order
                      col = c("#33608C", "#8867A1"),  # Colors for MM and AC respectively
                      warn = TRUE)

dend <- dend %>% 
  set("branches_lwd", 1.1)

# set("branches_k_color", value = c("#33608C", "#33608C", "#8867A1"), k = 3) 

par(
  mar = c(4, 1, 0, 1),
  mgp = c(1.3, 0.2, 0)
)

plot(dend, xlab = expression("log(CPM)"), leaflab = "none", 
     cex.lab = 0.7, cex.axis = 0.6, horiz = TRUE)

colored_bars(
  colors    = color_matrix, 
  dend      = dend,
  rowLabels = c("Species", "Condition"),
  cex.rowLabels = 0.6,
  y_shift   = 3.5,
  horiz = TRUE
)

# Close the PNG device
dev.off()
```
Distance between fibrotic zone from Acomys with Mus samples
```{r}
# Combine the dataset
combined_counts <- mm_fibrotic %>%
  inner_join(., mm_sham, by = join_by(GeneID)) %>%
  inner_join(., ac_fibrotic, by = join_by(GeneID))

meta_data <- get_meta(combined_counts)
```

Distance between sham zone from Acomys with Mus samples
```{r}
# Combine the dataset
combined_counts <- mm_fibrotic %>%
  inner_join(., mm_sham, by = join_by(GeneID)) %>%
  inner_join(., ac_sham, by = join_by(GeneID))

meta_data <- get_meta(combined_counts)
```

Legend
```{r}
library(ggplot2)
library(cowplot, quietly = TRUE)
```

Species legend
```{r}
# Create dummy data and plot
dummy_data <- data.frame(
  x = 1:2,
  y = 1:2,
  group = c("M. musculus", "A. cahirinus")
)

p <- ggplot(dummy_data, aes(x = x, y = y, color = group)) +
  geom_point(size = 3) +
  scale_color_manual(values = c(
    "M. musculus" = "#33608C", 
    "A. cahirinus" = "#8867A1"
  )) +
  theme_void() +
 theme(
    legend.position = "bottom",
    legend.direction = "vertical",
    legend.box.margin = margin(0, 0, 0, 0),
    legend.margin = margin(0, 0, 0, 0),
    legend.text = element_text(face = "italic")
  ) +
  labs(color = "Species")

# Extract legend using the correct position
legend <- get_plot_component(p, "guide-box-bottom")

# Create new plot with just the legend
g <- ggdraw(legend)

# Save the legend
ggsave("species_legend.png", g, width = 5, height = 0.8, bg = "white")
```

Zone legend
```{r}
# Create dummy data and plot
dummy_data <- data.frame(
  x = 1:2,
  y = 1:2,
  group = c("Fibrotic MI region", "Sham condition")
)

p <- ggplot(dummy_data, aes(x = x, y = y, color = group)) +
  geom_point(size = 3) +
  scale_color_manual(values = c(
    "Fibrotic MI region" = "#CA3B43",
    "Sham condition" = "#C7C9CA"
  )) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical",
    legend.box.margin = margin(0, 0, 0, 0),
    legend.margin = margin(0, 0, 0, 0)
  ) +
  labs(color = "Condition")

# Extract legend using the correct position
legend <- get_plot_component(p, "guide-box-bottom")

# Create new plot with just the legend
g <- ggdraw(legend)

# Save the legend
ggsave("condition_legend.png", g, width = 5, height = 0.8, bg = "white")
```
