---
title: "Tomoseq profile count plots"
author: "Arne Kukkonen"
date: "2025-01-15"
output: html_document
---

This script visualizes the counts of selected genes in profile across all slices, per conditon, per sample. The counts are normalized with the median-of-ratios method, across all samples within single species (Sham + MI14D condition). The genes are chosen as marker genes based on literature, to identify the 3 regions of post-MI remodeling (remote zone, border zone, infarct/fibrotic zone). 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(patchwork)
```

Directory paths
```{r}
script_input_dir <- "./input_datasets/"
marker_gene_dir <- paste0(script_input_dir, "marker_genes/")

script_output_dir <- "./output_dir/"
if (!dir.exists(script_output_dir)) {
  dir.create(script_output_dir)
}

rdata_output_dir <- "./plots_rdata/"
if (!dir.exists(rdata_output_dir)) {
  dir.create(rdata_output_dir)
}
```

Function for loading the normalized count data
```{r}
load_counts <- function(file_path) {
  counts <- readRDS(file_path) %>%
    as.data.frame() %>%
    round(., digits = 1) %>%
    rownames_to_column("Gene_EnsID")
  return(counts)
}
```

To do it with count data that has been normalized separately per-species
```{r}
mm_counts <- load_counts(paste0(script_input_dir, "MM_Tomoseq_counts_1-1orth_normalized.RData"))
ac_counts <- load_counts(paste0(script_input_dir, "AC_Tomoseq_counts_1-1orth_normalized.RData"))
```

Load the marker gene data
```{r}
# Function for loading the marker gene data 
load_markers <- function(file_path) {
  markers <- read_excel(file_path) %>%
    filter(!if_all(everything(), is.na), str_starts(Acomys_EnsID, "ENSACHG")) %>%
    select(-Gene_name, -Acomys_EnsID, -Source) %>%
    fill(Classification, .direction = "down")
    
}

bz_markers <- load_markers(paste0(marker_gene_dir, "border_zone_markers.xlsx"))
fz_markers <- load_markers(paste0(marker_gene_dir, "fibrotic_zone_markers.xlsx"))
cm_markers <- load_markers(paste0(marker_gene_dir, "cardiomyocyte_markers.xlsx"))
```

Filter the count data to only contain the required genes
```{r}
marker_genes <- c(bz_markers$Mus_EnsID, fz_markers$Mus_EnsID, cm_markers$Mus_EnsID)
mm_counts <- mm_counts %>% filter(Gene_EnsID %in% marker_genes)
ac_counts <- ac_counts %>% filter(Gene_EnsID %in% marker_genes)
rm(marker_genes)
```

Functions to create the plot data 
```{r}
# Function to get the sample numbers to define sample 1 and sample 2 on the plot. 
format_sample_numbers <- function(data, sample_condition) {
  data_fmt <- data %>%
    filter(Condition == sample_condition) %>%
    mutate(
      Replicate = if_else(Replicate == max(Replicate), 2, 1)
    )
  return(data_fmt)
}


prepare_plot_data <- function(count_data, marker_data) {
  plot_data <- count_data %>%
    inner_join(., marker_data, by = join_by(Gene_EnsID == Mus_EnsID)) %>%
    select(-Gene_EnsID, -Classification) %>%
    relocate(Gene_symbol) %>%
    pivot_longer(., cols = -Gene_symbol, names_to = "Sample_name", values_to = "Expression") %>%
    mutate(
      Species = str_extract(Sample_name, "^[A-Z]{2}"),
      Slice = as.numeric(str_extract(Sample_name, "(?<=\\.slice)\\d+")),
      Condition = str_extract(Sample_name, "(?<=_)[A-Z0-9]+(?=\\.)"),
      Replicate = str_extract(Sample_name, "(?<=\\.)\\d+(?=\\.)"),
      Replicate = as.numeric(Replicate)
      )
  
  # Transform the sample replicate numbers to either 1 or 2
  sham_fmt <- format_sample_numbers(plot_data, "SHAM")
  mi14d_fmt <- format_sample_numbers(plot_data, "MI14D")
  plot_data <- rbind(sham_fmt, mi14d_fmt)
  
  return(plot_data)

}
```

Function to create a plot for a single species counts (Sham + MI14D)
```{r}
profile_plot <- function(count_data, marker_data, y_max, top_plot = TRUE, plot_title) {
  # Prepare plot data
  plot_data <- prepare_plot_data(count_data = count_data, marker_data = marker_data)
  species <- if_else(unique(plot_data$Species)=="MM", "M. musculus", false = "A. cahirinus")
  
  plot_data <- plot_data %>%
    mutate(
      Replicate_label = if_else(
        Condition == "SHAM",
        "Sham replicates",                  # one label for all SHAM lines
        paste("MI day 14 rep.", sprintf("%02d", Replicate))  # label e.g. "MI day 14 rep. 01"
      )
    )
  
  # Create plot base
  p <- ggplot(
    plot_data,
    aes(
      x = Slice,
      y = Expression,
      color = Replicate_label,
      linetype = Replicate_label,
      group = interaction(Condition, Replicate)  # keep lines separate
    )
  )
  
  # Add the zone labels to top plot only
  if (top_plot) {
    p <- p + annotate("text",
            x = 16.2,
            y = y_max * 1.02,
            label = "Infarct zone",
            hjust = 0,
            fontface = "bold.italic",
            size = 6,
            color = "#FA8D76"
          ) +
          annotate("text",
            x = 11.2,
            y = y_max * 1.02,
            label = "Mus border zone",
            hjust = 0,
            fontface = "bold.italic",
            size = 6,
            color = "#F5A040"
          )
  }

  p <- p +
    # Add infarct zone highlight 
    annotate("rect", 
         xmin = 16, xmax = 24,
         ymin = -Inf, ymax = Inf,
         alpha = 0.5,  # Controls transparency
         fill = "#FEB1A3"
    ) +
    # Add border zone highlight 
    annotate("rect", 
         xmin = 11, xmax = 15,
         ymin = -Inf, ymax = Inf,
         alpha = 0.5,  # Controls transparency
         fill = "#FFC685"
    ) +
    
    # Add the sample count lines
    geom_line(aes(alpha = Replicate_label, linewidth = Replicate_label)) +
    
    # Set the color for each sample line
    scale_colour_manual(
      values = c(
        "MI day 14 rep. 01" = "#B57FA4",
        "MI day 14 rep. 02" = "#49A4B9",
        "Sham replicates"   = "gray50"
      ),
      name = "Samples"
    ) +
    # Set line width for every sample
    scale_linewidth_manual(
    values = c(
        "MI day 14 rep. 01" = 1.2,
        "MI day 14 rep. 02" = 1.2,
        "Sham replicates"   = 1.5
      ),
      guide = "none"  # Hide from legend if you don't want it shown
    ) +
    # Set line type for every sample
    scale_linetype_manual(
      values = c(
        "MI day 14 rep. 01" = "solid",
        "MI day 14 rep. 02" = "solid",
        "Sham replicates"   = "11"
      ),
      guide = "none"  # Hide the linetype legend
    ) +
    # Set line transparency for every sample
    scale_alpha_manual( 
      values = c(
        "MI day 14 rep. 01" = 1,
        "MI day 14 rep. 02" = 1,
        "Sham replicates"   = 0.8
      ),
      guide = "none"  # Hide alpha from legend
    ) +
    
    # Set the x-axis breaks to represent the Tomo-seq slices
    scale_x_continuous(breaks = seq(1, 24, 1), limits = c(1, 24), expand = expansion(add = c(0.3, 0.3))) +
    scale_y_continuous(limits = c(0, y_max*1.05)) +
    theme_bw() +
    labs(x = "Tomo-seq slice", y = "Normalized read count") +
    annotate("text",
      x = 1,  # position at first slice
      y = y_max * 1.04,  # position near top of plot
      label = species,
      hjust = 0,  # left-align text
      fontface = "bold.italic",
      size = 6,  # equivalent to 18pt
      color = "gray25"
    ) +
    theme(
      # Only show legend for the top plot
      legend.position = if_else(top_plot, "right", "none"),
      legend.title = element_text(size = 18, face = "bold"),  # Added: increase legend title size
      legend.text = element_text(size = 18),
      
      # Grid lines
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_line(linewidth = 1.1, color = "gray90"),
      
      # Show x axis text and label only for the bottom plot
      axis.title.x = if(top_plot) element_blank() else element_text(size = 18, vjust = -1),
      axis.text.x = if(top_plot) element_blank() else element_text(angle = 0, vjust = 0.6, size = 16, face = "bold"),
      axis.ticks.x = if(top_plot) element_blank() else element_line(),
      
      # Show y axis label only for the top plot
      # axis.title.y = if(top_plot) element_text(size = 18, vjust = 1, hjust = -0.2) else element_blank(),
      axis.title.y = element_blank(),
      axis.text.y = element_text(size = 14, face = "bold", colour = "grey40"),
      
      # Panel border
      panel.border = element_rect(color = "grey15", fill = NA, linewidth = 1),
      
      plot.title = element_text(size = 22, colour = "grey30", face = "bold")
      
      #plot.margin = margin(t = 15, r = 15, b = 15, l = 15, unit = "pt")
    )
  
  p <- p + guides(
    color = guide_legend(
        override.aes = list(linewidth = 2, linetype = c("solid", "solid", "11")),
        keywidth = unit(2, "cm"),
        keyheight = unit(1.5, "lines"),
    )
  )
  
  p <- p + ggtitle(if (!top_plot) NULL else plot_title)
  
  return(p)
}
```

Function to iteratively create profile plots for every marker in every category
```{r}
create_marker_plots <- function(mouse_counts, acomys_counts, marker_data, zone) {
  # Create an output directory for the zone 
  out_zone_dir <- paste0(script_output_dir, zone)
  dir.create(out_zone_dir, recursive = TRUE)
  
  # Define the marker classes for the zone 
  marker_classes <- unique(marker_data$Classification)
  
  # Iterate through all marker classes
  for (i in 1:length(marker_classes)) {
    
    # Define the current class
    input_class <- marker_classes[i]
    
    # Create the output directory for a single marker class
    out_dir <- paste0(out_zone_dir, "/", input_class, "/")
    dir.create(out_dir, recursive = TRUE)
    
    # Define the genes in a class
    genes_in_class <- marker_data %>%
      filter(Classification == input_class) %>%
      pull(Gene_symbol)
    
    # Iterate through the genes
    for (j in 1:length(genes_in_class)) {
      
      input_gene <- genes_in_class[j]
      
      # Subset the marker data for the input gene
      input_marker_data <- marker_data %>%
        filter(Gene_symbol == input_gene)
      
      # Define the maximum y value (count) to create a matching scale between both plots
      mus_max_count <- mouse_counts %>% filter(Gene_EnsID == input_marker_data$Mus_EnsID) %>%
        summarise(max_value = max(across(-Gene_EnsID))) %>%
        pull()
      acomys_max_count <- acomys_counts %>% filter(Gene_EnsID == input_marker_data$Mus_EnsID) %>%
        summarise(max_value = max(across(-Gene_EnsID))) %>%
        pull()
      max_count <- max(mus_max_count, acomys_max_count)
      
      # Create a plot for each species with the filtered marker data
      plot_mouse <- profile_plot(count_data = mouse_counts, marker_data = input_marker_data, 
                                 y_max = max_count, plot_title = input_gene)
      
      plot_acomys <- profile_plot(count_data = acomys_counts, marker_data = input_marker_data, 
                                  y_max = max_count, plot_title = input_gene, top_plot = FALSE)
      
      # Stack the plots on top of one another and add the marker gene symbol as title
      plot_combined <- plot_mouse / plot_acomys
      
      saveRDS(plot_combined, paste0(rdata_output_dir, input_gene, ".RData"))
      # If you want border around the combined plot, add this to theme()
      #plot.background = element_rect(color = "black", linewidth = 2, fill = NA)
      
      # save the plot
      out_path <- paste0(out_dir, input_gene, "_marker_profile.png")
      ggsave(plot = plot_combined, filename = out_path, width = 20, height = 9)
    }
  }
}
```

Create the profile plots for every zone
```{r}
create_marker_plots(mouse_counts = mm_counts, acomys_counts = ac_counts, marker_data = cm_markers, zone = "remote_zone")
create_marker_plots(mouse_counts = mm_counts, acomys_counts = ac_counts, marker_data = bz_markers, zone = "border_zone")
create_marker_plots(mouse_counts = mm_counts, acomys_counts = ac_counts, marker_data = fz_markers, zone = "infarct_zone")
```

### Select markers to combine into a single figure
```{r}
fz_selected <- c("Col1a1", "Col3a1", "Bgn", "Lum")
bz_selected <- c("Myh7", "Des", "Nppa", "Xirp2")
rz_selected <- c("Myh6", "Tnnc1", "Atp2a2", "Ryr2")
```

List the plot files for every category and combine under a single letter. Then combine these under a single figure name
```{r}
# Function to get the list of plot files of the selected markers
get_plot_list <- function(marker_vector) {
  plot_list <- list()
  for (i in 1:length(marker_vector)) {
    marker_gene <- marker_vector[i]
    file_path <- paste0(rdata_output_dir, marker_gene, ".RData")
    plot_file <- readRDS(file_path)
    plot_list[[marker_gene]] <- plot_file
  }
  return(plot_list)
}
```


Fibrotic zone figure
```{r}
fz_plot_list <- get_plot_list(fz_selected)
fz_figure <- wrap_plots(fz_plot_list, ncol = 2) +
  plot_layout(guides = 'collect') +
  plot_annotation(
    title = "Fibrotic zone markers",
    theme = theme(
      # Make the title larger and bold
      plot.title = element_text(size = 26, face = "bold", hjust = 0),
      # Add a black border around the combined figure
      plot.background = element_rect(
        color = "grey50", 
        linewidth = 1, 
        fill = NA
      )
    )
  )

# Save with adjusted dimensions for better visibility
ggsave("fz_figure.png", 
       fz_figure,
       width = 27,  # Increased width slightly
       height = 15, # Increased height to accommodate all plots
       units = "in",
       dpi = 100)
```

Border zone figure
```{r}
bz_plot_list <- get_plot_list(bz_selected)
bz_figure <- wrap_plots(bz_plot_list, ncol = 2) +
  plot_layout(guides = 'collect') +
  plot_annotation(
    title = "Border zone markers",
    theme = theme(
      # Make the title larger and bold
      plot.title = element_text(size = 26, face = "bold", hjust = 0),
      # Add a black border around the combined figure
      plot.background = element_rect(
        color = "grey50", 
        linewidth = 1, 
        fill = NA
      )
    )
  )

# Save with adjusted dimensions for better visibility
ggsave("bz_figure.png", 
       bz_figure,
       width = 27,  # Increased width slightly
       height = 15, # Increased height to accommodate all plots
       units = "in",
       dpi = 100)
```

Remote zone figure
```{r}
rz_plot_list <- get_plot_list(rz_selected)
rz_figure <- wrap_plots(rz_plot_list, ncol = 2) +
  plot_layout(guides = 'collect') +
  plot_annotation(
    title = "Remote zone markers",
    theme = theme(
      # Make the title larger and bold
      plot.title = element_text(size = 26, face = "bold", hjust = 0),
      # Add a black border around the combined figure
      plot.background = element_rect(
        color = "grey50", 
        linewidth = 1, 
        fill = NA
      )
    )
  )

# Save with adjusted dimensions for better visibility
ggsave("rz_figure.png", 
       rz_figure,
       width = 27,  # Increased width slightly
       height = 15, # Increased height to accommodate all plots
       units = "in",
       dpi = 100)
```

combine all
```{r}
# Combine the 3 sub-figures without losing each figure's own border & title
combined_figure <- 
  wrap_elements(full = fz_figure) /  # use '/' to stack vertically
  wrap_elements(full = bz_figure) /
  wrap_elements(full = rz_figure)

# (Optional) If you also want a grand title or other top-level annotations,
# you can add another plot_annotation() call here:
combined_figure <- combined_figure + 
  plot_annotation(
    caption = "Figure x",  # Using caption instead of title
    theme = theme(
      plot.caption = element_text(size = 30, face = "bold", hjust = 0)
    )
  )

# Finally, save
ggsave("all_zones.png", combined_figure, 
       width = 27, height = 35, units = "in", dpi = 100)
```